PoVaTiM: A Possibilistic Valid-Time Model \\ \vspace*{0.04truein}

### 2012-08-01 ijcis-pos-model/sources/intro.tex
preprint version

A very specific type of information systems are database systems. A database contains data representing real objects or concepts. In real world, some aspects or properties of objects are time-variant or time-related. e.g., the moment of a bank transaction and the status of an employee in a company, are time-related and time-variant notions, respectively.
A temporal database schema is a database schema that models objects with time-related or time-variant properties. However, the modelling of temporal aspects has a direct impact on the consistency of the temporal database, because the temporal nature of these aspects imposes extra integrity constraints and suitable ways of interaction with the human user.

For example, let us consider an employee's database containing data about the positions of employees in a company. Two dates are stored: the starting and ending dates respectively of the contract. It is clear that an employee cannot have a contract with an ending date without an starting date. Without further cautions, a manager in the company could insert a new contract for an employee who is already working for the company. A temporal database model will typically constrain record insertion and prevent similar modelling inconsistencies.

@@

Notably, in 1994, ``A Consensus Glossary of Temporal Database Concepts'' was published~\cite{Dyreson1994}. For this publication, 44 temporal database researchers, among them, some of the main researchers in this field, cooperated to reach a consensus on the nature and definitions of some of the main temporal database concepts and terminology. This glossary was updated in 1998~\cite{Dyreson1998}.

@@

An interesting issue in temporal modelling concerns relationships between temporal notions. In this sense, Allen~\cite{Allen1983} studied temporal relationships between time intervals (and as a special case time points). Among others, the querying of temporal databases has greatly profited from these temporal relationships, because they allowed more powerful user-specified temporal query demands, by allowing to express more complex relationships between the temporal notions in the temporal expressions in the query. For example, a query like `who were the department heads when Thomas worked for the institution' can be evaluated using operators similar to Allen's ones.

Humans handle temporal information using certain temporal notions like time intervals or time points~\cite{Dyreson1994}, and they often have to deal with imperfections like imprecision, vagueness, uncertainty or inconsistencies possibly contained in the descriptions of these temporal notions. These possible imperfections in descriptions of temporal notions determine an important issue in temporal modelling. Consider as an example the description of the temporal notion in a sentence like `The Belfry of Bruges was finished between 01/01/1201 A.D. and 31/12/1300 A.D.'. This sentence contains imperfection because of the uncertainty in the used time-related expression. It is known that the building was finished on a single day, but this day is not precisely known.

To allow information systems to cope with these and similar data imperfections, many approaches adopt fuzzy sets~\cite{Zadeh1965} for the representation and management of temporal information~\cite{Mitra1994,Nagypal2003,Billiet2011,Dubois2003}. The temporal relationships studied by Allen were fuzzified by several authors~\cite{Ohlbach2004,Nagypal2003,Schockaert2008}.   Garrido et al. ~\cite{Garrido2009} presented a compact representation for the time and defined different relationships among time intervals by using a combination of regular fuzzy comparisons. Also ~\cite{Garrido2009,Pons2011} studied uncertainty in temporal expressions concerning time intervals. Other approaches, like~\cite{Qiang2009}, use rough sets~\cite{Pawlak1995} to represent time intervals.

@@

In addition to temporal modelling, some attention has been paid to temporal reasoning~\cite{Allen1983}. Although temporal reasoning is not discussed in this paper, it should be noted that, among others, Dubois and Prade et al.~\cite{Dubois2003,DuBois1989} have dealt with fuzziness and uncertainty in temporal reasoning.

The present work defines and implement a model for properly represent and manage uncertainty in valid-time specification in a relational database. Our work is focused on both the proposal of an appropriate formal framework to suitably manage time in databases and the implementation of an DML that allows to the user the transparent use of this proposal. None of the previous research offer a database model to accomplish this task.

This way together with the theoretical model, we also present and explain the main operations of the manipulation language for a temporal database which stores the valid-time periods of the objects affected by imprecision. The rest of the work is organized as follows. Section \ref{sec:prelim} presents some background concepts about both possibility theory and temporal databases. In section \ref{sec:time-rep} the representation of the valid-time intervals in the database is explained. Section \ref{sec:temporal-model} explains the main concepts of the temporal Data Manipulation Language (DML) and its implementation. Finally, Section \ref{sec:conclusions} presents the conclusions and some guidelines for future research work.

>>>>

A very specific type of information systems are database systems. A database contains data representing real objects or concepts. In real world, some aspects or properties of objects are time-variant or time-related. E.g., the moment of a bank transaction and the status of an employee in a company, are time-related and time-variant notions, respectively.
A temporal database schema is a database schema that models objects with time-related or time-variant properties. However, the modelling of temporal aspects has a direct impact on the consistency of the temporal database, because the temporal nature of these aspects imposes extra integrity constraints and suitable ways of interaction with the human user.

% For example, let us consider an employee's database containing data about the positions of employees in a company. Two dates are stored: the starting and ending dates respectively of the contract. It is clear that an employee cannot have a contract with an ending date without an starting date. Without further cautions, a manager in the company could insert a new contract for an employee who is already working for the company. A temporal database model will typically constrain record insertion and prevent similar modelling inconsistencies.

For example, let us consider a library database and, concretely, the modelling of the presence of books in the library. Two dates are stored: the loan and the return date. It is clear that a book cannot be loaned again until it is returned. Without further cautions, a library employee could loan the same book several times even if it is not returned. A temporal database model will typically constrain record insertion and prevent similar modelling inconsistencies.

@@

%Notably, in 1994, ``A Consensus Glossary of Temporal Database Concepts'' was published~\cite{Dyreson1994}. For this publication, 44 temporal database researchers, among them, some of the main researchers in this field, cooperated to reach a consensus on the nature and definitions of some of the main temporal database concepts and terminology. This glossary was updated in 1998~\cite{Dyreson1998}.

@@

An interesting issue in temporal modelling concerns relationships between temporal notions. In this sense, Allen~\cite{Allen1983} studied temporal relationships between time intervals (and as a special case time points). Among others, the querying of temporal databases has greatly profited from these temporal relationships, because they allow more powerful user-specified temporal query demands, by allowing to express more complex relationships between the temporal notions in the temporal expressions in the query. For example, a query like `who were the department heads when Thomas worked for the institution' can be evaluated using operators similar to Allen's ones.

Humans handle temporal information using certain temporal notions like time intervals or time points~\cite{Dyreson1994}, and they often have to deal with imperfections like imprecision, vagueness, uncertainty or inconsistencies possibly contained in the descriptions of these temporal notions. These imperfections in descriptions of temporal notions determine an important issue in temporal modelling. Consider as an example the description of the temporal notion in a sentence like `The Belfry of Bruges was finished between 01/01/1201 A.D. and 31/12/1300 A.D.'. This sentence contains imperfection because of the uncertainty in the used time-related expression. It is known that the building was finished on a single day, but this day is not precisely known.

To allow information systems to cope with these and similar data imperfections, many approaches adopt fuzzy sets~\cite{Zadeh1965} for the representation and management of temporal information~\cite{Mitra1994,Nagypal2003,Billiet2011,Dubois1983,Dubois2003}. The temporal relationships studied by Allen were fuzzified by several authors~\cite{Ohlbach2004,Nagypal2003,Schockaert2008}.   Garrido et al. ~\cite{Garrido2009} presented a compact representation for the time and defined different relationships among time intervals by using a combination of regular fuzzy comparisons. Also ~\cite{Garrido2009,Pons2011} studied uncertainty in temporal expressions concerning time intervals. Other approaches, like~\cite{Qiang2009}, use rough sets~\cite{Pawlak1995} to represent time intervals.

@@

In addition to temporal modelling, some attention has been paid to temporal reasoning~\cite{Allen1983}. Although temporal reasoning is not discussed in this paper, it should be noted that, among others, Dubois and Prade et al.~\cite{Dubois2003,DuBois1989} have dealt with fuzziness and uncertainty in temporal reasoning. Finally, in \cite{rita1,rita2}, an approach to the linguistic summarization of data with time dimension is presented.

The present work defines and implements a model for properly representing and managing uncertainty in valid-time specification in a relational database. Our work is focused on both the proposal of an appropriate formal framework to suitably manage time in databases and the implementation of a DML that allows to the user the transparent use of this proposal. None of the previous research offers a database model to accomplish this task.

This way, together with the theoretical model, we also present and explain the main operations of the manipulation language for a temporal database which stores the valid-time periods of the objects affected by imprecision. The rest of the work is organized as follows. Section \ref{sec:prelim} presents some background concepts about both possibility theory and temporal databases. In section \ref{sec:time-rep} the representation of the valid-time intervals in the database is explained. Section \ref{sec:temporal-model} explains the main concepts of the temporal Data Manipulation Language (DML) and its implementation. Finally, Section \ref{sec:conclusions} presents the conclusions and some guidelines for future research work.
